```{r, echo=FALSE}
cat(paste("(C) (cc by-sa) Wouter van Atteveldt, file generated", format(Sys.Date(), format="%B %d %Y")))
```

> Note on the data used in this howto: 
> This data can be downloaded from http://piketty.pse.ens.fr/files/capital21c/en/xls/, 
> but the excel format is a bit difficult to parse at it is meant to be human readable, with multiple header rows etc. 
> For that reason, I've extracted csv files for some interesting tables that I've uploaded to 
> http://vanatteveldt.com/uploads/rcourse/data

Playing with data in R
========================================================


To demonstrate R, we will use the data from Piketty's 'Capital in the 21st Century' 

```{r}
download.file("http://vanatteveldt.com/wp-content/uploads/rcourse/income_topdecile.csv", 
              destfile="income_topdecile.csv")
income = read.csv("income_topdecile.csv")
```

We've downloaded a csv file and read it into a new variable `income`, which should appear in your environment list. 
You can click on the file to inspect it visually, but we can also use the `head` command:

```{r}
head(income, n=10)
```

As you can see, the values are NA (missing) for most rows, especially in the earlier period.
Let's throw out all data containing missing values using the `na.omit` function:

```{r}
income = na.omit(income)
head(income)
```

Much better. 
Now, we can list the variables in the file using `names` and get the numbers of rows or columns with `nrow` and `ncol`, respectively:

```{r}
names(income)
nrow(income)
ncol(income)
```

We can also ask for a summary of each of the variables in the file using the `summary` command:

```{r}
summary(income)
```

This lists the range, mean, etc. for each variable. 
We can select any column from a data frame using variable$column:

```{r}
income$U.S.
```

This gives a vector of numbers representing the different cells in that column. 
We can use various functions such as `mean`, `sum`, and `length` to get information about a vector.

```{r}
length(income$U.S.)
mean(income$U.S.)
mean(income$Europe)
```

As perhaps expected, the mean income inequality in Europe is lower than than in the U.S..
Let's do a t-test to see if the difference is significant:

```{r}
t.test(income$U.S., income$Europe, paired=T)
```

So, with p<.02 we can conclude that the income distribution in the U.S. is more unequal than in Europe. 
Let's make a simple plot of the income inequality in the U.S. and Europe
(reproducing fig 9.8 on page 324)

```{r}
plot(x=income$Year, y=income$U.S., type="l", ylab="Top decile income share", xlab="Year", ylim=c(0, 0.5))
lines(x=income$Year, y=income$Europe, col="red")
```

As you can see, income distribution in pre-WWI Europe is actually more unequal than in the U.S., 
but this is reversed during the 1910's and inequality diverges after the 1970's. 
Still, the lines are probably correlated:

```{r}
cor.test(income$U.S., income$Europe)
```

So, although the correlation is moderate at 0.43, it is not significant (due to a lack of data points)
