```{r, echo=FALSE}
cat(paste("(C) (cc by-sa) Wouter van Atteveldt, file generated", format(Sys.Date(), format="%B %d %Y")))
```

> Note on the data used in this howto: 
> This data can be downloaded from http://piketty.pse.ens.fr/files/capital21c/en/xls/, 
> but the excel format is a bit difficult to parse at it is meant to be human readable, with multiple header rows etc. 
> For that reason, I've extracted csv files for some interesting tables that I've uploaded to 
> https://github.com/vanatteveldt/learningr/tree/master/data.
> If you're accessing this tutorial from the githup project, these files should be in your 'data' sub folder automatically.

Merging and transforming data in R
==================================

Let's have a closer look at the relation at the accumulation of capital in various countries.
In chapter 5, separate values are introduced for private and public capital: 

```{r}
private.capital = read.csv("data/private_capital.csv")
public.capital = read.csv("data/public_capital.csv")
```

Combining data using cbind
----

First, let's combine the two data sets into a single set. 
This is done using `cbind`, which binds multiple columns (vectors or data frames) into a single data frame. 
It only works if the data  is alinged correctly and has the same amount of rows, so first we check this:

```{r}
nrow(private.capital)
nrow(public.capital)
table(private.capital$Year == public.capital$Year)
```

So they both have 41 rows and in all rows the Year variables are identical. 
Thus, we can use cbind here, omitting the first column of public.capital. 

```{r}
capital = cbind(private.capital, public.capital[-1])
head(capital)
```

Renaming variables
----

As could be seen in the `head` output, the column names are identical for both data sets, which causes a problem.
So, let's prepend the last 10 column names with `public.` using the `paste` function, which pastes two texts together.

```{r}
colnames(capital)[11:19] = paste("public", colnames(capital[11:19]), sep=".")
head(capital)
```

/Note that using `merge` rather than `cbind` would be more natural and would prevent these problems, but it is good to know how cbind and renaming works in any case/

In order to rename variables more easily in a 'manual' fashion, the `rename` command from the `plyr` package can be used. 

Note, if you get an error about missing a package, you can install it using the install.packages function (or you can use the 'Packages' panel in the bottom right in RStudio):

```{r, eval=FALSE}
install.packages("plyr")  # only needed the first time
```

Now, we can use `library(plyr)` to load the package, and use the rename command:

```{r}
library(plyr)
capital = rename(capital, c("U.K."="UK", "U.S."="US", "public.U.K."="public.UK", "public.U.S."="public.US"))
head(capital)
```

Variables to cases using melt
----

Now assume that we want to calculate the total capital (private plus public) per country.
One way to do that would be to simply sum the relevant parts of the data frame:

```{r}
total.capital = capital[, 2:10] + capital[, 11:19]
head(total.capital)
```

However, a more general way that will afford more flexibility later is to first transform both datasets from wide into long formats
(e.g. convert the variables to cases), and then merge the two sets. 
The `reshape2` package has a function `melt` that transforms variables to cases based on a list of index variables that identify the cases.

```{r}
library(reshape2)
public.long = melt(public.capital, id.vars="Year")
private.long = melt(private.capital, id.vars="Year")
head(public.long)
```

As you can see, we now have one row per year per country. 
We can now use `merge` to merge the two data frames. 
By default, `merge` merges data on identical column names. 
Since the `value` column is also shared between the data frames, 
and we don't want to merge on that column,
we need to specify the `by=` argument:

```{r}
capital = merge(public.long, private.long, by=c("Year", "variable"))
head(capital)
```

So now we have our data ready. 
Let's rename the column names to something more sensible, and then compute the total capital. 
We also save the file for later use using the `save` command.

```{r}
colnames(capital) = c("Year", "Country", "Public", "Private")
capital$Total = capital$Private + capital$Public
save(capital, file="capital.rdata")
```

Variables to cases
---

The counterparg of melting is called casting.
The `dcast` function works by using a 'casting formula' `rows ~ columns` to specify which variables to place in the rows and columns of the resulting data frame.

```{r}
d = dcast(capital, Year ~ Country, value.var="Total")
head(d)
```

So now we have a data frame that we can plot, e.g. using a for loop as above. 
`cast` can also work to make aggregate functions.
Suppose we would want to make a plot per decade rather than per year, we first make a new 'decade' value:

```{r}
capital$Decade = floor(capital$Year / 10) * 10
```

A good way to check whether a recode such as this succeeded is to tabulate decade and year:

```{r}
table(capital$Decade, capital$Year)
```

Now, we use the new Decade variable in the casting formula, and specify an aggregation function `mean`:

```{r}
d = dcast(capital, Decade ~ Country, value.var="Total", fun.aggregate=mean)
head(d)
```

Aggregate vs. cast
----

Instead of `cast`, we can also use the `normal' `aggregate` command. Cast and aggregate have slightly different applications. 
Aggregate allows multiple columns to be aggregated in one command (but all using the same aggregation function), 
but produces a 'long' data format, while `cast` can only aggregate a single variable, but can directly create a tabular data format.
For example, the following aggregates the public, private and total capital per decade and country:

```{r}
aggregated = aggregate(capital[c("Public", "Private", "Total")], by=capital[c("Decade", "Country")], FUN=mean)
head(aggregated)
```

*Note*: the `by=` argument in aggregate needs to be a list or data frame. Since `capital[, "Decade"]` returns a vector rather than a data frame, you must use either `capital["Decade"]` (ommitting the comma) or `capital[, "Decade", drop=F]`. For example, the first attempt below to aggregate the recent values per country fails because of this problem:

```{r,error=TRUE}
aggregated = aggregate(capital["Total"], by=capital[, "Country"], FUN=mean)
aggregated = aggregate(capital["Total"], by=capital[, "Country", drop=F], FUN=mean)
head(aggregated)
```

