```{r, echo=FALSE}
cat(paste("(C) (cc by-sa) Wouter van Atteveldt, file generated", format(Sys.Date(), format="%B %d %Y")))
```

R programming: Functions, conditions, and loops
====

All the code we've considered so far was executed from top to bottom as a list of instructions.
Sometimes, it can be useful to control the 'flow of execution', e.g. to repeat some statements or to only execute statements in certain conditions.

Conditional execution: if statements
---

A first flow control tool is the if statement. 
The if statement tests a condition, and only executes the conditional statements if it is true.
You can also provide an optional alternative statement which is executed if the condition is false.
The general syntax of if statements is as follows:

```
if (condition) {
  ...statements...
} else {
  ... statements
}
```

Where the part from `else` may be omitted. 
Note that the curly braces {} are used here to group multiple statements. 
If you use a single statements you can omit the braces, but it never hurts to use them.

A possible application of such a statement is to only execute a long-running task (such as getting data from an external data base)
if we don't have the data in a file already.
This can be achieved for checking for the file name and only creating the data if the saved version cannot be found:

```{r}
if (file.exists("cached_data.rda")) {
  message("Loading data from cache")
  load("cached_data.rda")
} else {
  message("Retrieving data from Internet")
  library(foreign)
  popdata = read.dta("http://www.ats.ucla.edu/stat/stata/examples/mlm_ma_hox/popular.dta")
  save(popdata, file="cached_data.rda")
}
```

The first time you execute these statements, it should indicate that it is getting the data from the Internet, and should create the `cached_data.rda` file.
If you execute the statements again, it will load the data from that file instead of retrieving it from the Internet again.

*Note*: Note the difference between the if statement and the `ifelse` function: 
The if statement expects a single value for the condition, and will execute a piece of code depending on that value.
By contrast, ifelse works on whole vectors, returning a list of values drawn from the 'if' and the 'else' depending on the condition.

Repetition: for loops
----

In some cases, you want to run the same operation a number of times.
For example, you might want to repeat a data cleaning operation on a number of different columns.
Consider the income inequality data we used earlier:

```{r}
income = read.csv('data/income_toppercentile.csv')
head(income, 10)
```

As seen earlier, the data contains a lot of missing values.
Although this might not be sensible for time series data, suppose we would want to replace missing values with the column means.
We can do this relatively easily for a single column:

```{r}
income$Canada[is.na(income$Canada)] = mean(income$Canada, na.rm = T)
head(income)
```

However, if we have a lot of columns it can be tedious to repeat this line for every column,
and that also creates the risk of making a mistake for one of the columns.
Using a for loop we can repeat an operation for multiple cases, in this case the various columns.
In a for loop, there is always a looping variable (or index variable) that iterates over a set of values.
The general syntax is like this:

```
for (index in values) {
  ... statements ...
}
```

The statements within the curly braces are then repeated for each value in values, 
and in each repetition the looping variable (here named 'index') will be one of those values.

As a simple example, let's just print the column names of all columns except the first:

```{r}
for (col in colnames(income)[-1]) {
  print(col)
}
```

So, to replace the NA's by the column mean for all columns, we plug the 'Canada' statement above into the loop,
replacing Canada by the variable `col`:

```{r}
for (col in colnames(income)[-1]) {
  income[[col]][is.na(income[[col]])] = mean(income[[col]], na.rm = T)
}
head(income)
```

Writing your own Functions
----

As we saw earlier, most work in R is done through functions, ranging from simple functions like `mean()` to complex functions like `lm()` and `plot()`.
In R, you can also define your own functions using a syntax like shown below:

```
functionname = function(arguments) {
  ... statements ...
  return(value)
}
```

The `arguments` allow you to operate the function on specific values or with specific options.
For example, above we called `mean(income$Canada, na.rm=T)`. 
In that call, `mean` is the name of the function, and `income$Canada` and `na.rm=T` are the two arguments that we 'pass' to the function.
Finally, a function can return a value, either with an explicit `return` call such as shown above, 
or if such a call is ommitted the last value is returned.

So, we could define a function that counts the number of non-missing values in a vector:

```{r}
n.valid = function(values) {
  return(sum(!is.na(values)))
}
```

Now we can call this function with different vectors like this:

```{r}
n.valid(c(1,2,3))
n.valid(c(NA, NA, 1, 2, NA))
```

Since our new function works just like existing functions, we can use it in various places.

For example, we might want to create a table of non-missing values per decade and per country.
To do this, lets first reload the income data, melt it, and create a 'decade' column:

```{r}
library(reshape2)
income = read.csv("data/income_toppercentile.csv")
long = melt(income, id.vars = "Year")
long$decade = floor(long$Year / 10) * 10
head(long)
```

As we saw earlier, if we cast data with multiple observations per call, we can specify the function used to aggregate these observations, e.g. to get the mean:

```{r}
dcast(long, decade ~ variable, fun.aggregate = mean)
```

Now, rather than using the built-in mean function, we can use our own `n.valid` function to get the number of non-missing observations:

```{r}
dcast(long, decade ~ variable, fun.aggregate = n.valid)
```

Like for loops, functions are useful if you want to run a number of statements multiple times.
They can also be combined, for example by turning the column mean imputation shown above into a function:

```{r}
impute.mean = function(values) {
   values[is.na(values)] = mean(values, na.rm=T)
   return(values)
}
```

And we can use this function e.g. to impute the missing values for Canada:

```{r}
impute.mean(income$Canada)
```

Now that we have the impute.mean function, we can use that in our for loop:

```{r}
income = read.csv("data/income_toppercentile.csv")
for (col in colnames(income)[-1]) {
  income[[col]] = impute.mean(income[[col]])
}
head(income)
```

As a final note, similar to if and for statements, the curly braces can be ommitted if the function only contains a single statement, so the following gives the same result:

```{r}
impute.mean = function(values) ifelse(is.na(values), mean(values, na.rm=T), values)
impute.mean(income$Canada)
```
