---
title: "Using AmCAT from R"
author: "Wouter van Atteveldt"
date: "May 25, 2016"
output: pdf_document
---

```{r, echo=F}
head = function(...) knitr::kable(utils::head(...))
```

This handout describes how to connect to AmCAT from R to conduct queries, download metadata, and upload articles and create new article sets.
You will need an account on a working AmCAT server, you can create a free account at [https://amcat.nl](https://amcat.nl) or install your own server,
see [https://github.com/amcat/amcat](the AmCAT github page).

You will also need to install the `amcatr` package from github, for which you will need the devtools package:

```{r, eval=F}
install.packages("devtools")
devtools::install_github("amcat/amcat-r")
```

# Connecting to AmCAT

On every computer you need to save your AmCAT password once::

```{r, eval=F}
library(amcatr)
amcat.save.password("https://amcat.nl", "your_username", "your_password")
```
```{r, echo=F, message=F}
library(amcatr)
```

Next, you can connect using the `amcat.connect` function, storing the connection details in an object `conn`.
The token that this creates is valid for 24 hours, so you need to run this command every session:

```{r}
conn = amcat.connect("https://amcat.nl")
```

# Retrieving article (meta)data

The `amcat.getarticlemeta` command allows you to retrieve the metadata from an article set.
Using the columns keyword, you can specify which columns to select, e.g. headline, medium, and author.

```{r, message=F}
meta = amcat.getarticlemeta(conn, 41, 29454, dateparts = T, columns=c("medium", "date"))
head(meta)
```

# Querying AmCAT

You can use the `amcat.hits` function to run a keyword query which returns the number of hits per document for a query:

```{r, message=F}
h = amcat.hits(conn, c("mortgage*", "greek* OR greece*"), labels=c("mortgage", "greece"), sets=29454)
head(h)
```

You can supply multiple queries, and can also supply labels for the queries with the labels argument.

While the hits command returns a row per document per query, you can also get aggregate data directly with the `amcat.aggregate` command,
specifying the aggregation axes such as medium, date, and keyword:

```{r, message=F}
t = amcat.aggregate(conn, "mortgage*", sets=29454, axis1="year", axis2="medium")
head(t)
```

We can also do the aggregation within R by merging the hits data with the meta data and using the `aggregate` command.
For example, this code will create a plot of hits per newspaper per week:

```{r}
h = merge(meta, h)
perweek = aggregate(h["count"], h[c("year", "query")], sum)
library(ggplot2)
ggplot(perweek, aes(x=year, y=count, color=query)) + geom_line()
```

# Uploading articles

It is also possible to upload articles from R using the `amcat.upload.articles` command.
This allows you to e.g. upload articles form a csv file or folder, or retrieve articles from an API and upload them.

```{r}
d = data.frame(headline=c("A test", "Another test"), date=as.Date(c("2001-01-01", "2008-12-31")))
setid = amcat.upload.articles(conn, project=1, articleset = "Test set",
  medium="test", headline=d$headline, date=d$date, text=d$headline)
arts = amcat.getarticlemeta(conn, project = 1, articleset=setid, columns = c("date", "headline"))
head(arts)
```

# Adding articles to article sets


Finally, you can add existing articles to a new or existing article set. 
For example, we could add the first articles from the set we used earlier to our new test set:

```{r}
articles = meta$id[1:3]
amcat.add.articles.to.set(conn, project = 1, articleset = setid, articles=articles)
```

By specifying an articleset.name rather than an existing set id, we can also create a new articleset from a selection of articles:


```{r}
articles = meta$id[1:3]
setid2 = amcat.add.articles.to.set(conn, project = 1, articleset.name="New test set", articles=articles)
```

