---
title: "Getting Sentiment Resources from the Internet"
author: "Wouter van Atteveldt"
date: "June 3, 2016"
output: pdf_document
---

```{r, echo=F}
head = function(...) knitr::kable(utils::head(...))
```
This handout describes how to download and parse a sentiment lexicon and collection of reviews from the Internet. 

You don't need to run this as the results are saved in the github repository and can be downloaded directly: [lexicon](rawgit.com/vanatteveldt/learningr/master/data/lexicon.rds); [reviews](rawgit.com/vanatteveldt/learningr/master/data/reviews.rds).
However, it can be interesting to see how to download files and parse the 'custom' file formats in R.

Pittsburgh Sentiment Lexicon
===

There are many sentiment dictionaries available for download.
For this handout, we use a dictionary developed at the University of Pittsburgh that can be freely downloaded from http://mpqa.cs.pitt.edu/lexicons/subj_lexicon/.

```{r, message=F}
url = "http://mpqa.cs.pitt.edu/data/subjectivity_clues_hltemnlp05.zip"
file = "subjectivity_clues_hltemnlp05/subjclueslen1-HLTEMNLP05.tff"
download.file(url, destfile="lexicon.zip")
unzip("lexicon.zip", file=file)
lines = scan(file, what = "", sep="\n")
head(lines)
```

The file contained is in a somewhat strange format, with one word per line coded as name=value pairs.
So, we make a 'read_pairs' function that we apply to the result of extracting all name=value pairs per line:

```{r}
read_pairs = function(x, fields) { 
  x = x[x[,2] %in% fields, ]
  values = x[,3]
  names(values) = x[,2]
  values
}
m = stringr::str_match_all(lines, "(\\w+)=([^=]+)(?= |$)")
fields = m[[1]][,2]
lexicon = plyr::ldply(m, read_pairs, fields=fields)
saveRDS(lexicon, "data/lexicon.rds")
head(lexicon)
```

Amazon Reviews
===

For this exercise we will download the Amazon reviews in the 'automotive' category as published at http://jmcauley.ucsd.edu/data/amazon/.

These reviews are stored in a gzipped while which contains one json record per line, 
so we use scan to split the file into lines, and then read each line with a custom functions that converts the line from json and into a data frame.
We then use `ldply` to apply this function to each line:

```{r}
url = "http://snap.stanford.edu/data/amazon/productGraph/categoryFiles/reviews_Automotive_5.json.gz"
download.file(url, destfile="reviews.json.gz")
lines = scan(gzfile("reviews.json.gz"), sep = "\n", what="")
readline = function(line) {
  x = rjson::fromJSON(line)
  x$helpful = NULL
  as.data.frame(x)    
}
reviews = plyr::ldply(lines, readline)
saveRDS(reviews, "data/reviews.rds")
head(reviews)
```
