Machine Learning using RTextTools
=================================

_(C) 2014 Wouter van Atteveldt, license: [CC-BY-SA]_

Machine Learning or automatic text classification is a set of techniques to train a statistical model on a set of annotated (coded) training texts, that can then be used to predict the category or class of new texts.

R has a number of different packages for various machine learning algorithm such as maximum entropy modeling, neural networks, and support vector machines. 
`RTextTools` provides an easy way to access a large number of these algorithms.

In principle, like 'classical' statistical models, machine learning uses a number of (independent) variables called _features_ to predict a target (dependent) category or class. 
In text mining, the independent variables are generally the term frequencies, and as such the input for the machine learning is the document-term matrix. 

Training models using RTextTools
----

So, the first step is to create a document-term matrix. 
We only want to use the documents for which the sentiment is known (for now).
As before, we use the `achmea.csv` that can be downloaded from [github](https://raw.githubusercontent.com/vanatteveldt/learningr/master/achmea.csv).

```{r}
library(RTextTools)
d = read.csv("data/reviews.csv")
ds = d[!is.na(d$SENTIMENT), ]
m = create_matrix(ds$CONTENT, language="dutch", stemWords=F)
```

The next step is to create the RTextTools _container_. 
This contains both the d-t matrix and the manually coded classes,
and you specify which parts to use for training and which for testing.

To make sure that we get a random sample of documents for training and testing,
we sample 80% of the set for training and the remainder for testing.
(Note that it is important to sort the indices as otherwise GLMNET will fail)

```{r}
n = nrow(m)
train = sort(sample(1:n, n*.8))
test = sort(setdiff(1:n, train))
length(train)
length(test)
```

Now, we are ready to create the container:

```{r}
c = create_container(m, ds$SENTIMENT, trainSize=train, testSize=test, virgin=F)
```

Using this container, we can train different models:

```{r}
SVM <- train_model(c,"SVM")
MAXENT <- train_model(c,"MAXENT")
GLMNET <- train_model(c,"GLMNET")
```

Testing model performance
----

Using the same container, we can classify the 'test' dataset

```{r}
SVM_CLASSIFY <- classify_model(c, SVM)
MAXENT_CLASSIFY <- classify_model(c, MAXENT)
GLMNET_CLASSIFY <- classify_model(c, GLMNET)
```

Let's have a look at what these classifications yield:

```{r}
head(SVM_CLASSIFY)
nrow(SVM_CLASSIFY)
```

For each document in the test set, the predicted label and probability are given. 
We can compare these predictions to the correct classes manually:

```{r}
t = table(SVM_CLASSIFY$SVM_LABEL, as.character(ds$SENTIMENT[test]))
t
```

(Note that the as.character cast is necessary to align the rows and columns)
And compute the accuracy:

```{r}
sum(diag(t)) / sum(t)
```


Analytics
----

To make it easier to compute the relevant metrics, RTextTools has a built-in analytics function:

```{r}
analytics <- create_analytics(c, cbind(SVM_CLASSIFY, GLMNET_CLASSIFY, MAXENT_CLASSIFY))
names(attributes(analytics))
```

The `algorithm_summary` gives the performance of the various algorithms, with precision, recall, and f-score given per algorithm:

```{r}
analytics@algorithm_summary
```

The `label_summary` gives the performance per label (class):

```{r}
analytics@label_summary
```

Finally, the `ensemble_summary` gives an indication of how performance changes based on the amount of classifiers that agree on the classification:

```{r}
analytics@ensemble_summary
```

The last attribute, `document_summary`, contains the classifications of the various algorithms per document, and also lists how many agree and whether the consensus and the highest probability classifier where correct:

```{r}
head(analytics@document_summary)
```

Classifying new material
----

New material (called 'virgin data' in RTextTools) can be coded by placing the old and new material in a single container. 
We now set all documents with a sentiment score as training material, and specify `virgin=T` to indicate that we don't have coded classes on the test material:

```{r}
m_full = create_matrix(d$CONTENT, language="dutch", stemWords=F)
coded = which(!is.na(d$SENTIMENT))
c_full = create_container(m_full, d$SENTIMENT, trainSize=coded, virgin=T)
```

We can now build and test the model as before:

```{r}
SVM <- train_model(c_full,"SVM")
MAXENT <- train_model(c_full,"MAXENT")
GLMNET <- train_model(c_full,"GLMNET")
SVM_CLASSIFY <- classify_model(c_full, SVM)
MAXENT_CLASSIFY <- classify_model(c_full, MAXENT)
GLMNET_CLASSIFY <- classify_model(c_full, GLMNET)
analytics <- create_analytics(c_full, cbind(SVM_CLASSIFY, GLMNET_CLASSIFY, MAXENT_CLASSIFY))
names(attributes(analytics))
```

As you can see, the analytics now only has the `label_summary` and `document_summary`:

```{r}
analytics@label_summary
head(analytics@document_summary)
```

The label summary now only contains an overview of how many where coded using consensus and probability. The document_summary lists the output of all algorithms, and the consensus and probability code. 